---
title: "Gov 50 Final Project"
author: "Ezra Sharpe"
description: "Final project"
output:
  distill::distill_article:
    self_contained: false
---

```{r}
library(tidyverse)
library(dplyr)
library(readr)
library(ggplot2)

```

## Introduction

What are the key relationships which structure European states’ dependency on 
Gazprom’s natural gas exports?

This study will unpack this question using an originally made ‘Gazprom 
Dependency’ dataset, which will facilitate an analysis of what kinds of 
variables influence the relationships Gazprom has with European states. Within 
this broad umbrella question, there are also a range of sub-questions which this
project aims to confront; does reliance on Gazprom natural gas vary between FSU 
countries and non-FSU countries? Over which years are these relationships most 
pronounced? Does reliance on Gazprom natural gas vary between NATO and non-NATO
member states? What are the general trends in changing natural gas production 
levels over time, and what events could be potential drivers of this change? 

Studying Gazprom in close detail is an important research agenda, not least 
because energy has been the main denominator in European states’ (in)ability to
respond to Russian aggressions in the 20th century. For instance, Stern’s (2005)
‘The Future of Russian Gas and Gazprom’ outlined almost two decades ago the 
importance of understanding energy dependency in Europe – and this certainly has
a feeling of foresight given how events transpired ever since in 2014 and 2022.
The research in this context is mainly interesting because of the diversity of 
variables available to analyse within the novel dataset and the broad temporal 
spectrum, with data ranging from 1991 to 2020. Overall, the primary aim of this
study is to understand which political, economic, and militaristic factors 
underpin the different relationships European states have with Gazprom, and why 
these trends may have changed over time. In doing so, we might be able to better
understand how European states can strive for increased energy independence in
the future and divest themselves from Gazprom and the Russian state. 

My null hypothesis, or H(0), is that domestic natural gas production levels, 
military alliances, and corruption levels are not statistically significant 
factors structuring the relationship between Gazprom and client states (further
discussion of significance levels to come). By posing this as a null hypothesis,
if any one of these relationships are found to be statistically significant, we
can reject the null hypothesis and conduct further analysis on causality. With
regards to the relative significance of each variable, I would hypothesize that
domestic natural gas production rates have the most statistically significant 
inverse relationship with Gazprom reliance. That is to say that as domestic 
natural gas production increases, I would predict some sort of proportional 
decrease in reliance on Gazprom. 


```{r}

gazprom_data <- read_csv("Gazprom_data.csv")

gazprom_data <- subset(gazprom_data, select = c(Year, Country, Production, Gazprom_Imports, Natural_Gas_Reliance, Total_Reliance, Difference_in_Reliance, Geographic_Status, Military_Alliance, Control_of_Corruption))

gazprom_data$Natural_Gas_Reliance <- as.numeric(sub("%", "", gazprom_data$Natural_Gas_Reliance))/100
gazprom_data$Total_Reliance <- as.numeric(sub("%", "", gazprom_data$Total_Reliance))/100
gazprom_data$Difference_in_Reliance <- as.numeric(sub("%", "", gazprom_data$Difference_in_Reliance))/100

view(gazprom_data)
```

