---
title: "Gov 50 Final Project"
author: "Ezra Sharpe"
description: "Final project"
output:
  distill::distill_article:
    self_contained: false
---

```{r}
library(tidyverse)
library(dplyr)
library(readr)
library(ggplot2)

```

## Milestone 2 - dataset proposal and hypothesis

What is the relationship between the number of diplomatic contacts between 
countries, the volume of trade they do, and the likelihood of going to war with 
each other? This study will unpack this question using three different datasets
from the ‘correlates of war’ collection: the Diplomatic Exchange data set, the 
Trade dataset, and the Militarized Interstate Disputes dataset. The codebook for
each outlines that a consistent notation of state ID numbers are used across all
the datasets, meaning that I will be able to merge all 3 into a single dataset.
In doing so, I plan to explore how the number of diplomatic contacts between 
states influences both the volume of trade and the likelihood of interstate war
between recognised countries. I will not have to filter out countries with no 
disputes by virtue of the fact that the interstate war dataset does not measure
conflict in binary terms (0 = no war, 1 = war), but rather measures it through a spectrum of ‘threats’, ‘display of force’, and ‘application of force’. This 
therefore allows me to measure the relationship between diplomatic personnel and
the likelihood of war, even in cases where full-scale kinetic warfare did not 
occur. My initial hypothesis, or H(0), is that as the number of diplomatic 
personnel and contacts increases between states, so does the volume of trade, 
and the likelihood of war falls accordingly. This is because diplomacy is 
usually considered to be a sign of mutual recognition, and disputes between 
states necessitates that either country does not recognise some aspect of the 
other. However, my alternative hypothesis (H1) suggests that there could be a 
slight ‘survivors bias’ to this case, and that an increase in diplomatic 
personnel between states may actually be an indication of deteriorating 
relations. In this sense, as diplomacy ramps up, perhaps the likelihood of war
does too. Overall, I aim to understand through this study whether diplomacy has historically served as a useful peacekeeping tool, or whether it is perhaps an 
outdated practice – and in which case we ought to consider new political means 
to maintain peace. 

## Milestone 3

Apologies, my data has changed significantly because the above was too complicated.
I tried and it was too much of a headache. We're now looking at a dataset on foreign gifts donated to US educational institutions. We will be looking at how gifts vary by donors, geographies, size, and school. 

The first visualisation I want to make should be a reflection of the size of the top 20 total foreign donations recieved by schools in the US.

```{r}


top_20_schools <- Foreign_Gifts_All_gifts_dataset |>
  rename(School_Name = `School Name`, Donation_Amount = `Amount`) |>
  group_by(School_Name) |>
  summarize(Total_Donations = sum(as.numeric(gsub('[$,]', '', Donation_Amount))) / 1000000000) %>%
  arrange(desc(Total_Donations)) %>%
  top_n(20)

top_20_schools

```

This is now the top 20 total recipients expressed in billions. Let's make a graph.

```{r}

bar_chart <- ggplot(top_20_schools, aes(x = reorder(School_Name, -Total_Donations), y = Total_Donations)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Top 20 Schools by Total Donation Amount (in Billions USD)",
       x = "School Name",
       y = "Total Donation Amount (Billions USD)") +
  theme(axis.text.x = element_text(angle = 70, hjust = 1))

print(bar_chart)

```



